% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\documentclass[
]{article}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}

\usepackage{hyperref}
\usepackage{ulem}  % For underlining links

\hypersetup{
    colorlinks=false,         % Disable colored links (boxes will appear)
    citebordercolor=red,      % Border color for citation links
    linkbordercolor=red,      % Border color for internal links
    urlbordercolor=blue,      % Border color for external links (URLs)
    pdfborder={0 0 2}         % Specifies the border thickness: {horizontal vertical thickness}
}


% Only underline external links (URLs)
\let\oldhref\href
\renewcommand{\href}[2]{\ifx#1\urlprefix\oldhref{#1}{#2}\else\uline{\oldhref{#1}{#2}}\fi}


\renewcommand{\[}{\begin{equation}}
\renewcommand{\]}{\end{equation}}


\setcounter{secnumdepth}{5}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\usepackage[]{biblatex}
\addbibresource{refs.bib}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}

\title{Deep strong coupling}
  \author{Matt Lilley}
  \date{\today}  % Default to today if no date is provided

\begin{document}
\maketitle

\section{Introduction}\label{introduction}

Coupling refers to the interaction between two systems, where the total
energy is not a simple sum of the energies of each system. Instead, the
total energy also depends on the combined states of both, with each
system influencing the other in ways that cannot be separated. We often
express the ideas through a Lagrangian or Hamiltonian.

The aim of these notes is to build some intuition for a quantum systems
that have extremely strong coupling. In the
\href{https://www.nature.com/articles/s42254-018-0006-2}{quantum optics
literature}, the use the terms:

\begin{itemize}
\tightlist
\item
  Weak
\item
  Strong
\item
  Ultra strong
\item
  Deep strong
\end{itemize}

to describe the different regimes.

We'll begin with a classical example and use the quantum optics language
above. We choose the example of coupled pendulums because it turns out
each quantum state with a well defined energy behaves like it's own
pendulum (see
e.g.~\href{https://journals.aps.org/pra/abstract/10.1103/PhysRevA.85.052111}{Briggs
et.al}).

\section{Coupled pendulums}\label{coupled-pendulums}

For example, consider two identical pendulums of length \(l\) and mass
\(m\) connected by a spring whose stiffness is characterised by \(k\).
In the small angle approximation (\(\theta_1, \theta_1 \ll 1\)), the
Hamiltonian is:

\[
H_{\rm pen} = \frac{m l^2}{2} \dot{\theta}_1^2 + \frac{m g l}{2} \theta_1^2 + \frac{m l^2}{2} \dot{\theta}_2^2 + \frac{m g l}{2} \theta_2^2 + \frac{1}{2} k l^2(\theta_1 - \theta_2)^2
\]

The first four terms are the simple sum of each individual pendulum. The
third term arrises due to the coupling.

More abstractly, we can write:

\[
H = H_1 + H_2+ V_{\text{coupling}}
\]

and associate frequencies to the different parts:

\begin{itemize}
\tightlist
\item
  \(\omega_1 = \sqrt{g/l}\)
\item
  \(\omega_2 = \sqrt{g/l}\)
\item
  \(\omega_{\rm coupling} = \sqrt{k/m}\)
\end{itemize}

The system behaves quite differently depending on the strength of the
coupling which is proportional to the different frequencies.

\subsection{Weak coupling}\label{weak-coupling}

In reality, there is always dissipation which cannot be properly
captured in a Hamiltonian description. We can however define a
dissipation rate \(\gamma_{\rm diss}\) whose magnitude also changes the
system behaviour.

When
\(\omega_{\rm coupling} \ll \gamma_{\rm diss} \ll \omega_1,\omega_2\),
the coupling is described as weak.

There are very small changes in the natural frequencies of the system as
compared to the uncoupled case:

\begin{itemize}
\tightlist
\item
  \(\omega_1 \rightarrow \omega_+ = \sqrt{g/l}\)
\item
  \(\omega_2 \rightarrow \omega_- = \sqrt{g/l + 2k/m}\)
\end{itemize}

The energy does not however move back and forwards between the pendulums
because of the large dissipation.

\subsection{Strong coupling}\label{strong-coupling}

When the coupling is ``strong'' in the sense that
\(\gamma_{\rm diss} \ll \omega_{\rm coupling} \ll \omega_1,\omega_2\),
dissipation is small enough to allow energy to be slowly
\href{https://www.youtube.com/watch?v=CjJVBvDNxcE&t=57s}{exchanged
between the two pendulums}. The motion is characterised by individual
swings happening with a frequency \(\approx \sqrt{g/l}\) where the
amplitude of those swings gradually undulates on a timescale
characterised by \(\omega_{\rm exchange} = k/2m\omega_p\). This exchange
happens most effectively when the pendulums have the same length, so
that their natural frequencies are the same .

Strong coupling allows us to still conceptually consider the pendulums
as having well defined identities in the sense that they have their own
natural frequencies. As the coupling becomes larger, this is no longer
the case.

\subsection{Ultra strong coupling}\label{ultra-strong-coupling}

When the coupling is ``ultra strong'' in the sense that
\(\gamma_{\rm diss} \ll \omega_{\rm coupling} \sim 0.1\times  \omega_1,\omega_2\),
energy exchange happens on the time scale of a single swing of one of
the pendulums. The two natural natural frequencies can be noticeably
discerned, \(\omega_+ = \sqrt{g/l}\) when both pendulums move ``in
phase'' (the spring is not stretched) with one another and
\(\omega_- = \sqrt{g/l + 2k/m}\) when the pendulums move ``out of
phase''.

The coupling is getting strong enough so that more energy can be
exchanged between pendulums of different lengths.

This exact boundary for this regime is somewhat artificial - there is
nothing particularly special about the value
\(0.1 \omega_{\rm 1}, \omega_2\). This value was first used as part of
the quantum optics literature.

\subsection{Deep strong coupling}\label{deep-strong-coupling}

When
\(\gamma_{\rm diss} \ll \omega_1,\omega_2 \lesssim \omega_{\rm coupling}\)
, the coupling begins to dominate over everything else and we enter into
a regime called ``Deep strong coupling''. Energy transfer between the
two pendulums is so fast that it's almost instantaneous and so it's not
possible to move one pendulum without the other - they act as a single
rigid body.

\section{Rabi model}\label{rabi-model}

A canonical quantum example is a single two level system (TLS)
interacting with a quantised field often called the Rabi model. The
Hamiltonian can be written as:

\[
H_{\rm Rabi} = \frac{\Delta E}{2} \sigma_z + \hbar\omega\left(a^{\dagger}a +\frac{1}{2}\right) + U\left( a^{\dagger} + a \right)(\sigma_+ + \sigma_-)
\label{eq:rabiH}
\]

where \(\Delta E\) is the transition energy between the 2 levels of the
TLS, \(\hbar\omega\) is the energy of each quantum of the field, and
\(U\) is the coupling constant between the TLS and the field. The
\(\sigma\) operators are the Pauli spin matrices and \(a^{\dagger}\),
\(a\) are the field creation and annihilation operators respectively.

It's worth noting that we're using the Pauli spin matrices as a
mathematical tool to describe two levels. Just keep in mind that we're
not really talking about spin angular momentum here.

Although a TLS has just 2 states (denoted \(|\pm\rangle\)), the
quantised field has infinitely many states (denoted by the number of
quanta \(|n\rangle\)). The combined state of the system (denoted
\(|n, \pm\rangle\)) therefore has infinitely many states and so
conceptually the system behaves like infinitely many pendulums coupled
together. The frequency of these conceptual pendulums is determined by
the energy of the states.

Much like the classical example, the dynamics depend on the relative
sizes of the different terms in the Hamiltonian. For the quantum case
however, it's not enough just to compare the various constants
\(U, \hbar \omega, \Delta E\), we must also consider how many field
quanta \(n\) we have. This is because of how the field operators work:

\[
a^{\dagger} |n,\pm\rangle = \sqrt{n+1}|n+1,\pm\rangle \\
\]

\[
a |n,\pm\rangle = \sqrt{n}|n-1,\pm\rangle \\
\]

\[
a^{\dagger}a |n,\pm\rangle = n|n,\pm\rangle
\]

The more field quanta we have, the larger the field and coupling terms
will be compared to the TLS term.

\subsection{Weak coupling}\label{weak-coupling-1}

Much like the classical pendulums, the quantum states can suffer various
forms of ``dissipation''. As the quantum systems interacts with outside
systems, it can cause:

\begin{itemize}
\tightlist
\item
  Dephasing - where the phase relationship between each of the quantum
  states starts to change over time
\item
  Decoherence - the system is forced out of a superposition state and
  into a well defined state aka ``collapse of the wavefunciton''
\end{itemize}

If we define \(\hbar\gamma_{\rm diss}\) as a characteristic energy
associated with the above processes, then
\(\sqrt{n}U \ll \hbar \gamma_{\rm diss} \ll \Delta E , n\hbar \omega\)
defines the weak coupling regime. Spontaneous emission is the most
characteristic feature of weak coupling where an excited TLS with less
field quanta (e.g.~\(|n,+\rangle\)) is coupled to a ground state TLS
with more field quanta (e.g.~\(|n+1,-\rangle\) ). The coupling is
however not so strong that the field quanta can get reabsorbed.

\begin{quote}
Note that truly irreversible spontaneous emission also relies on there
being a continuum of states instead of discrete ones.
\end{quote}

\subsection{Strong coupling}\label{strong-coupling-1}

As in the classical case, when the coupling is strong in the sense that
\(\hbar\gamma_{\rm diss} \ll \sqrt{n}U \ll \Delta E , \hbar \omega\),
there is time for slow exchange between the different quantum states
(remember each state is like it's own pendulum). Unlike the classical
case, where it's energy that's exchanged, in the quantum case it's state
occupation probability \(|\psi|^2\) that's exchanged.

In order for exchange to occur effectively, the quantum states need to
have the same energy. This is equivalent to the conceptual pendulums
having the same length. Two such states are often described as
``resonant'' with one another. Whether or not the system has any
resonances depends on the relationship between \(\hbar \omega\) and
\(\Delta E\).

\subsubsection{Matched field and TLS}\label{matched-field-and-tls}

When the field is matched to the TLS, \(\Delta E = \hbar\omega\). This
is the most widely discussed regime in which a transition of the TLS
(often an atomic transition) results in the emission of a single field
quantum (often a photon). In a cavity where discrete states can be
arranged and field quanta can be confined, this results in occupation
probability oscillating between states like \(|n,+\rangle\) and
\(|n+1,-\rangle\). These oscillations are called
\href{https://en.wikipedia.org/wiki/Jaynes\%E2\%80\%93Cummings_model\#Vacuum_Rabi_oscillations}{Rabi
oscillations} which have a frequency
\(\Omega/\hbar\omega \sim \sqrt{n}U/\hbar\omega\)

\subsubsection{Mismatched field and TLS}\label{mismatched-field-and-tls}

When the field is mismatched to the TLS, \(\Delta E \neq \hbar\omega\) .
If the mismatch is arbitrary, e.g.~\(\Delta E/ \hbar\omega = 2.83677\)
then Rabi oscillations cannot occur because the coupling term
(\(\sim \sqrt{n}U\)) is still very small compared to \(\Delta E\) and
\(\hbar\omega\) and so it cannot accommodate any energy mismatch.

If however, \(\Delta E = m\hbar \omega\) where \(m=3, 5, 7 ...\) then
\(|n,+\rangle\) is resonant with, \(|n+3,-\rangle\), \(|n+5,-\rangle\),
\(|n+7,-\rangle ...\) and so Rabi oscillations can once again occur. The
frequency is slower
\(\Omega/\hbar\omega \sim (\sqrt{n}U/\hbar\omega)^m\) and so for larger
\(m\) the emission of multiple quanta becomes less and less likely.

\subsection{Ultra strong coupling}\label{ultra-strong-coupling-1}

When the coupling becomes a sizeable fraction of the TLS and field
quantum,
\(\hbar\gamma_{\rm diss} \ll \sqrt{n}U \sim 0.1 \times \Delta E , \hbar \omega\),
non-resonant states begin to gain significant occupancy. For example, a
system can start out in state \(|0,+\rangle\) with 100\% probability and
overtime a state \(|1,+\rangle\) can gain a non-trivial amount of
occupation probability. Although this superficially appears to violate
energy conservation, the energy in the coupling is no longer small and
so all terms in the Hamiltonian need to be considered when thinking
about energy conservation.

The coupling term can also accommodate energy mismatches between the TLS
and the oscillator, e.g.~\(\Delta E/ \hbar\omega = 2.83677\) vs
\(\Delta E/ \hbar\omega = 3\). This makes it easier to observe the
emission of multiple quanta.

\subsection{Deep strong coupling}\label{deep-strong-coupling-1}

When the coupling becomes on the same order or greater than the TLS and
field quantum
\(\hbar\gamma_{\rm diss} \ll \Delta E , \hbar \omega \lesssim \sqrt{n}U\),
then TLS transitions and creation/annihilation of field quanta can no
longer be understood by simply thinking about the TLS and field
exchanging energy with each other and the coupling as a kind of glue
between the two. The coupling term has an ``identity'' all of its own.

This regime was first theoretically explored in
\href{https://journals.aps.org/prl/abstract/10.1103/PhysRevLett.105.263603}{2010
by Casanova at.al} where a simpler definition of ``deep strong
coupling'' was given as:

\[
\frac{U}{\hbar\omega} \gtrsim 1
\label{eq:deepstrongcoupling}
\]

Indeed, if their condition is satisfied then
\(\hbar \omega \lesssim \sqrt{n}U\) is guaranteed.

Let's first consider the case (as Casanova did) that the TLS energy is
small in the sense that \(\Delta E < \hbar \omega\). If the coupling is
in the deep storing regime so that \(U/\hbar\omega \gtrsim 1\) then,
from an energy conservation point of view, the coupling term can
spontaneously create field quanta. When a quanta gets created, then
\(n\) increases which means the coupling term \(\sim \sqrt{n}U\)
increases which means more quanta can be made. We can figure out how
many can be made by equating the field energy to the coupling energy in
the Hamiltonian:

\[
n\hbar\omega \sim \sqrt{n}U
\label{eq:couplingbalancefield}
\]

This gives us:

\[
n \sim \left(\frac{U}{\hbar\omega}\right)^2
\label{eq:selfconsistentn}
\]

When you work out the detailed maths, you end up with
\(n = 4(U/\hbar\omega)^2\).

When the TLS energy is not small (\(\Delta E \gtrsim \hbar \omega\))
then there is an additional energy equivalence to consider:

\[
\Delta E \sim \sqrt{n}U
\label{eq:couplingbalancetls}
\]

Eq. \(\ref{eq:couplingbalancefield}\) and Eq.
\(\ref{eq:couplingbalancetls}\) can be solved simultaneously to
eliminate \(n\). This gives us a relationship between
\(U,\Delta E,\hbar\omega\):

\[
\begin{aligned}
\Delta E &\sim \sqrt{\left(\frac{U}{\hbar\omega}\right)^2}U \\
\Delta E &\sim \frac{U^2}{\hbar\omega} \\
U &\sim \sqrt{\Delta E \hbar\omega} \\
\frac{U}{\hbar\omega} &\sim \sqrt{\frac{\Delta E}{\hbar\omega}}
\end{aligned}
\label{eq:uconstraint}
\]

For the case when the TLS energy dominates over the field,
\(\Delta E \gg \hbar \omega\), Eq. \(\ref{eq:uconstraint}\) is a more
appropriate coupling threshold to consider than Eq.
\(\ref{eq:deepstrongcoupling}\). This regime is sometimes called
``dispersive deep strong coupling'' as was first coined by
\href{https://journals.aps.org/pra/abstract/10.1103/PhysRevA.95.013827}{Felicetti
et.al in 2017}. Again, when you do the detailed maths, you get an extra
constant so that:

\[
\frac{U}{\hbar\omega} \gtrsim \frac{1}{2}\sqrt{\frac{\Delta E}{\hbar\omega}}
\label{eq:superradianttransition}
\]

In the extreme case when \(\Delta E/\hbar\omega \rightarrow \infty\)
then Eq. \(\ref{eq:superradianttransition}\) represents a boundary of
what's called a superradiant phase transition (detailed in
\href{https://journals.aps.org/prl/abstract/10.1103/PhysRevLett.115.180404}{2015
by Hwang et.al}). When you go above this critical coupling, the system
undergoes a phase change where the lowest energy state involves a
non-zero amount of field quanta. In other words, above this threshold
the TLS freely exchanges energy with the field and the usual
restrictions around having a matched TLS and field are not important.
Superradiant phase transitions have been discussed for much longer times
\href{https://en.wikipedia.org/wiki/Dicke_model\#Superradiant_transition_and_Dicke_superradiance}{in
relation to the Dicke model} and we'll come back to look at this later
on.

\subsubsection{Relativistic phonon nuclear
coupling}\label{relativistic-phonon-nuclear-coupling}

For acoustic phonons, we choose a notation \(\omega \equiv \omega_A\).
From notes on
\href{https://github.com/project-ida/notes/blob/main/pdf/Coupling\%20constants\%20in\%20nuclear\%20physics.pdf}{Coupling
constants in nuclear physics}, we derived the relativistic phonon
nuclear coupling as:

\[
\frac{U}{\hbar \omega_A} = \sqrt{\frac{2}{N}} \sqrt{\frac{\Delta E}{M c^2}} \sqrt{\frac{\Delta E}{\hbar \omega_A}} \times 10^{-3}
\label{eq:phononcoupling}
\]

where \(N\) is the number of nuclei involved in the phonon motion, and
\(M\) is the mass of the nucleus.

Typically, we imagine a phonon as being the quantised oscillatory motion
of many nuclei in a lattice. However, it's possible to arrange systems
in which the motion of an isolated nucleus is considered (see
e.g.~\href{https://www.nature.com/articles/s41467-021-21425-8}{Cat et.al
2021}) - in which case \(N=1\).

Our example Hamiltonian in Eq. \(\ref{eq:rabiH}\) is for a single TLS,
so we'll consider Eq. \(\ref{eq:phononcoupling}\) with \(N=1\). We'll
extend the Hamiltonian to many TLS later on.

For nuclear transitions mediated by phonons, \(\Delta E \sim MeV\)and
\(\hbar\omega_A \sim 10neV\) and so \(\Delta E/\hbar\omega \gg 1\).
Therefore, Eq. \(\ref{eq:superradianttransition}\) is the appropriate
superradiant threshold condition.

Substituting the expression for coupling (Eq.
\(\ref{eq:phononcoupling}\)) into the critical coupling expression (Eq.
\(\ref{eq:superradianttransition}\)) gives the following condition:

\[
\begin{aligned}
2\frac{U}{\hbar\omega_A}\sqrt{\frac{\hbar\omega_A}{\Delta E}} &\gtrsim 1 \\
2\sqrt{2} \sqrt{\frac{\Delta E}{M c^2}}  \times 10^{-3} &\ge 1
\end{aligned}
\label{eq:criticalphononcouplingexplicit}
\]

For nuclear transitions \(\Delta E \sim MeV\) and for a single nucleon
\(Mc^2 \sim GeV\) so it's already clear that for any sized nucleus we
won't enter the superradiant regime. For the sake of completeness, let's
evaluate Eq. \(\ref{eq:criticalphononcouplingexplicit}\) for a fusion
transition with \(\Delta E \approx 24MeV\) and for a palladium nucleus
with \(Mc^2 \approx 100GeV\)

\[
2\sqrt{2} \sqrt{\frac{24\times10^6}{10^{11}}}  \times 10^{-3} \approx 4 \times 10^{-5} \ll 1
\label{eq:criticalphononcouplingexplicitnumbers}
\]

This confirms that a single nucleus cannot become superradiant using
relativistic phonon nuclear coupling. Let's look at another type of
coupling that's also associated with oscillatory phonon motion.

\subsubsection{Electric dipole coupling}\label{electric-dipole-coupling}

For electric dipole coupling associated with an oscillating electric
field driving phonon motion, we continue to use the notation
\(\omega \equiv \omega_A\). From notes on
\href{https://github.com/project-ida/notes/blob/main/pdf/Coupling\%20constants\%20in\%20nuclear\%20physics.pdf}{Coupling
constants in nuclear physics}, we derived the electric dipole phonon
coupling as:

\[
\frac{U}{\hbar \omega_A} = \frac{2\pi\sqrt{2}}{Z \sqrt{N}} \sqrt{\frac{M c^2}{\hbar \omega_A}} \frac{\hbar \omega_A}{E_L} A^{1/3} \times 6 \times 10^{-3}
\label{eq:dipolecoupling}
\]

If we once again take \(N=1\), thens substituting the expression for
coupling (Eq. \(\ref{eq:dipolecoupling}\)) into the critical coupling
expression (Eq. \(\ref{eq:superradianttransition}\)) gives the following
condition:

\[
\begin{aligned}
2\frac{U}{\hbar\omega_A}\sqrt{\frac{\hbar\omega_A}{\Delta E}} &\gtrsim 1 \\
\frac{4\pi\sqrt{2}}{Z} \sqrt{\frac{M c^2}{\Delta E}} \frac{\hbar \omega_A}{E_L} A^{1/3} \times 6 \times 10^{-3} &\gtrsim 1
\end{aligned}
\label{eq:criticaldipolecouplingexplicit}
\]

For palladium nuclear transitions mediated by acoustic phonons:

\begin{itemize}
\tightlist
\item
  \(A \approx 106\)
\item
  \(Z \approx 106\)
\item
  \(M c^2 \approx 10^{11}\) eV
\item
  \(\Delta E \approx 24 \times 10^{6}\) eV\\
\item
  \(\hbar \omega_A \approx 10^{-8}\) eV
\end{itemize}

The localization energy \(E_L\) is:

\[
E_L = \frac{\hbar c}{R_0} = \frac{6.6 \times 10^{-34} \times 3 \times 10^8}{10^{-15}}
\]

\[
= 2 \times 10^{-10} \text{ J} = 1.2 \times 10^9 \text{ eV} \approx 10^9 \text{ eV}
\]

And so evaluating Eq. \(\ref{eq:criticaldipolecouplingexplicit}\) gives:

\[
\frac{4 \pi\sqrt{2} }{106} \times \sqrt{\frac{10^{11}}{24\times10^{6}}} \times \frac{10^{-8}}{10^9} \times 6 \times 10^{-3} \times 106^{1/3} \approx 3\times 10^{-18} \ll 1
\]

And so we're even further away from the superradiant regime when
considering the electric dipole coupling associated with the phonon
motion.

We've so far looked at the Rabi model where the number of TLS is
\(N=1\). How does the story change when we have many TLS?

\subsection{Dicke model}\label{dicke-model}

When we have \(N\) identical TLS coupled to a single mode (i.e.~single
frequency) of a quantised field then Hamiltonian is called the Dicke
Hamiltonian. It's a simple extension of the Rabi Hamiltonian in Eq.
\(\ref{eq:rabiH}\) in the sense that we just add \(N\) copies of the TLS
terms as seen below:

\[
H_{\text{Dicke}} = \frac{\Delta E}{2} \sum_{i=1}^N \sigma_z^{(i)} +  \hbar\omega\left(a^{\dagger}a +\frac{1}{2}\right) + U \sum_{i=1}^N (a^\dagger + a) (\sigma_+^{(i)} + \sigma_-^{(i)})
\label{eq:dickeH}
\]

The states of this system are described by
\(|n, \pm, \pm, \pm, \pm, ... \rangle\).

It's worth emphasising that there is no spatial dependence of the field
in Eq. \(\ref{eq:dickeH}\). One way to understand this physically is
that all the TLS are very close together in the sense that they are
located in a region of space that is much smaller than the wavelength of
the mode. In that situation, all the TLS will experience the same
strength of field at any given moment - in other words the field appears
constant in space. This is the how Dicke originally presented his ideas
in \href{https://journals.aps.org/pr/abstract/10.1103/PhysRev.93.99}{his
1954 paper}. It's also possible to use this Hamiltonian to describe many
TLS arranged in a very special way so that they are placed at integer
multiples of the mode wavelength.

What we really want to understand is how the extra TLS are going to
affect the coupling term. Right now, Eq. \(\ref{eq:dickeH}\) is not in a
form that allows us to readily see this. We're going to rewrite it by
appealing to the physics of spin.

We noted earlier that the use of the Pauli spin matrices is just a
mathematical tool to describe two levels. Although we're not describing
spin here, we are working with what is often called ``pseudo spin''. The
\href{https://en.wikipedia.org/wiki/Angular_momentum_operator}{rules of
angular momentum} work just as well for pseudo angular momentum. In
particular, the rules of
\href{https://www.tcm.phy.cam.ac.uk/~bds10/aqp/lec7_compressed.pdf}{angular
momentum addition} and conservation.

This means that we can treat all the TLS together as if they are a
single object with many levels whose energies are determined by the
addition rules of pseudo angular momentum. The Hamilton then looks like:

\[
H =  \Delta E J_{z} + \hbar\omega\left(a^{\dagger}a +\frac{1}{2}\right) + 2U\left( a^{\dagger} + a \right)(J_{+} + J_{-})
\label{eq:dickeHpseudo}
\]

where the total pseudo total angular momentum operators (\(J\)) for
\(N\) TLS are:

\[
J_{+} + J_{-} = J_{x} = \frac{1}{2}\overset{N}{\underset{i=1}{\Sigma}} \sigma_{i x} \,\,\,\,\,\, J_{z} = \frac{1}{2}\overset{N}{\underset{i=1}{\Sigma}} \sigma_{i z}
\]

and noting that \(i\) in \(\sigma_i\) means that this operator only acts
on TLS number \(i\) .

When written in this way, each state can now be described in terms of 3
numbers \(|n, j, m\rangle\) where \(j\) describes the total pseudo
angular momentum number (which is conserved) and \(m\) describes the z
component of the total pseudo angular momentum (which can change). This
notation allows us to conveniently describe situations where excitations
are ``delocalised'' among the TLS. By far the most significant kind of
delocalised states are called ``Dicke states'' which have the largest
\(j=j_{\max} = N/2\). Dicke states are symmetric in the sense that if
you swap any of the TLS around, the state remains unchanged. For
example, consider a single excitation in 4 TLS - the Dicke state looks
like:

\[
\Psi_0 = \frac{1}{\sqrt{4}}\left(| 0, +, -, -, - \rangle + | 0, -, +, -, - \rangle + | 0, -, -, +, - \rangle + | 0, -, -, -, + \rangle \right)
\]

Notice that if you swap any two TLS, the state looks the same.

The above state can instead be described by \(j_{\max}= 4/2  = 2\) and
\(m = 1\times 1/2 + 3\times -1/2 =-1\)

\[
\Psi_0 = |0,2,-1>
\]

Dicke states with \(j=j_{\max}\) are significant because of the
acceleration properties that these Dicke states offer; something people
often describe as superradiance. We can start to get a sense of where
the super comes from by looking at the action of the
\href{https://en.wikipedia.org/wiki/Ladder_operator\#Angular_momentum}{ladder
operators} \(J_{+}\) and \(J_{-}\) which excite and de-excite the TLS.
This causes a raising and lowering of the \(m\) value like this:

\[
J_+ |n, j, m\rangle  =  \sqrt{j(j + 1) - m(m + 1)} |n, j, m + 1\rangle
\]

\[
J_- |n, j, m\rangle =  \sqrt{j(j + 1) - m(m - 1)} |n, j, m - 1\rangle
\]

These ladder operators are conceptually similar to the creation and
annihilation operators of the field. The details are however more
complicated due to the addition rules of angular momentum.

For the case with \textbf{all TLS excited} , \(m=N/2\): \[
J_- |n, N/2, N/2\rangle  = \sqrt{N} |n, N/2, N/2 - 1\rangle
\]

For the first de-excitation, the coupling terms gets enhanced by
\(\sqrt{N}\) . This might not seem surprising at first glance because we
have \(N\) TLS excited and so we should expect the rate of emission
(which go as the square of the coupling) to be enhanced by \(N\).

For the case of \textbf{a single excitation} , \(m=-N/2 + 1\):

\[
J_- |n, N/2, -N/2 +1\rangle  = \sqrt{N}|n, N/2, -N/2\rangle
\]

For this singe de-excitation the coupling term also gets enhanced by
\(\sqrt{N}\) . This is more surprising because the rate of emission
(which go as the square of the coupling) to be enhanced by \(N\) even
though there is only a single excitation.

For the case of \textbf{50\% excitation}, \(m=0\):

\[
J_- |n, N/2, 0 \rangle  = \sqrt{N^2 + N}|n, N/2, -1\rangle
\]

For the first de-excitation, the coupling terms gets enhanced by
\(\sim\sqrt{N^2}\) for large \(N\). In other words, the rate of emission
(which go as the square of the coupling) to be enhanced by \(N^2\) -
this is where the super in superradiance comes from.

It's should be noted that the effect of Dicke superradiance and
superradiant phase transitions are
\href{https://en.wikipedia.org/wiki/Dicke_model\#Superradiant_transition_and_Dicke_superradiance}{not
the same}. The former involves a transient enhancement in emission of
\(N\) TLS which ultimately ends up with all the TLS in their ground
state and the field quanta escape to infinity. The latter involves a
permanent change in the ground state a cavity system in which field and
TLS are both confined.

If we look back at the couplings in Eqs. \(\ref{eq:phononcoupling}\) and
\(\ref{eq:dipolecoupling}\), we can see a \(1/\sqrt{N}\) term appears to
reduce the coupling significantly for very large numbers of nuclei.
However, if we are able to take advantage of Dicke effects, then the
situation is very different:

\begin{itemize}
\tightlist
\item
  For fully excited systems, Dicke enhancement of the coupling scales
  like \(\sqrt{N}\) and so coupling for \(N\) nuclei is the same as for
  a single nuclei (from \(\sqrt{N}/\sqrt{N}\))
\item
  For half excited systems, Dicke enhancement of the coupling scales
  like \(N\) and so coupling for \(N\) nuclei scales like \(\sqrt{N}\)
  (from \(N/\sqrt{N}\))
\end{itemize}

\subsubsection{Deep strong coupling}\label{deep-strong-coupling-2}

We now understand that the coupling operators in Eq.
\(\ref{eq:dickeHpseudo}\) will produce terms like:

\begin{itemize}
\tightlist
\item
  \(\sqrt{n}\) from the field part
\item
  At least \(\sqrt{N}\) and at most \(N\) from the TLS part due to Dicke
  effects
\end{itemize}

If we take the most conservative Dicke enhancement then the condition
for deep strong coupling can be arrived at by simply replacing \(U\) in
Eq. \(\ref{eq:superradianttransition}\) with \(U\sqrt{N{}}\): \[
\frac{U\sqrt{N}}{\hbar\omega} \gtrsim \frac{1}{2}\sqrt{\frac{\Delta E}{\hbar\omega}}
\label{eq:dickesuperradianttransition}
\] As \(N\rightarrow \infty\) this condition triggers a
\href{https://royalsocietypublishing.org/doi/10.1098/rsta.2010.0333}{superradiant
phase transition} similar to what we saw in the Rabi model when
\(\Delta E / \hbar\omega \rightarrow \infty\).

If we look back at the couplings in Eqs. \(\ref{eq:phononcoupling}\) and
\(\ref{eq:dipolecoupling}\), we can see a \(1/\sqrt{N}\) terms will
cancel with the \(\sqrt{N}\) in Eq.
\(\ref{eq:dickesuperradianttransition}\). This means that our earlier
calculations with \(N=1\) will apply to an arbitrary number of TLS and
so we won't get closer to a superradiant phase transition by having more
TLS involved.

If however, we could use the most optimistic Dicke enhancement, then
we'd instead have:

\[
\frac{UN}{\hbar\omega} \gtrsim \frac{1}{2}\sqrt{\frac{\Delta E}{\hbar\omega}}
\label{eq:dickesuperradianttransition2}
\]

And so for relativistic phonon nuclear coupling (Eq.
\(\ref{eq:phononcoupling}\)) we'd have:

\[
2\sqrt{2}\sqrt{N} \sqrt{\frac{\Delta E}{M c^2}}  \times 10^{-3} \ge 1
\]

Using the same numbers as before in Eq.
\(\ref{eq:criticalphononcouplingexplicitnumbers}\) then:

\[
2\sqrt{2} \sqrt{N}\sqrt{\frac{24\times10^6}{10^{11}}}  \times 10^{-3} \approx 4 \times 10^{-5} \sqrt{N} \ge 1
\]

This gives us a condition on the number of nuclei that we need:

\[
N \gtrsim 8 \times 10^8
\]

If we were instead to consider a different transition, e.g.~the
\(\rm 14keV\) transition of \(\rm ^{57}Fe\) then:

\[
2\sqrt{2} \sqrt{N}\sqrt{\frac{14\times10^3}{5\times 10^{10}}}  \times 10^{-3} \approx 1.5 \times 10^{-6} \sqrt{N} \ge 1
\]

Which would gives us the following condition on the number of nuclei
that we need:

\[
N \gtrsim 4 \times 10^{11}
\]

\printbibliography


\end{document}
